Trees

### 2019-01-28 Trees_making.Rmd
added schematic trees examples and tree grafting

@@

A few things to note:  
1. We can use a tree with or without branch lengths (divergence measures).  
2. The tree has to be binary (strictly bifurcating = no polytomies, i.e. only two branches are coming from each node).  
3. The tree needs to be rooted (has a node representing the most recent common ancestor of all the species in the tree).   
4. The tree tip labels (representing species) must match **exactly** with the species names in the data set (otherwise MA models will not run with them).  
5. Trees can be also made and used for higher taxonomic levels than species, e.g. genus-level (not covered in this tutorial, but general principles are the same).

The general flow of tree-making work is shown in the diagram below.  
You will usually start with species list from the meta-analytic data set (use binomial Latin names for the species list and in your database). First, check if there is any phylogenetic tree available that already contains all your species (e.g. a super-tree). Currently, the easiest way to build a custom phylogenetic tree usable for meta-analysis is by using an R package called *rotl*, which accesses a synthetic super-tree from Open Tree of Life database (https://opentree.wikispaces.com/). We will use this approach in our examples below. Most of the time *rotl* will work well, especially with the common species. There are other super-trees available, e.g. for birds - we will use one of these as our example too. Overall, if you can find a tree that contains all your species, you just trim (subset) it down to your species. If a few species is missing, it is sometimes possible to substitute them with closely related species. If no existing phylogeny is readily usable, there are a few non-exclusive options depending on the range of species in the data set:  
 - If the species are very diverged (e.g. mixture of vertebrates and invertebrates), use NCBI Taxonomy Common Tree tool (http://www.ncbi.nlm.nih.gov/Taxonomy/CommonTree/wwwcmt.cgi). At the webpage, use an input box at the top of the page to search each species name or upload list of the species names from a file. This allows to generate a scaffold for a tree.  
 - If few major clades are present, assemble together sub-trees created from bigger trees, if these exist for these clades.    
 - Use phyloGenerator (http://willpearse.github.io/phyloGenerator/) or other phylogenetic software to create phylogeny based on DNA sequences, if a shared sequence set exists for a given set of species (check if the resulting phylogeny makes sense).

@@

@@

Extract and clean the list of species included. Note that genus and species name are stored in separate columns.

@@

@@

So, we run into a problem with one of the Populus names - need to fix this by using an alternative name.

```{r fix Curtis1998 species_Curtis1998 list}
populus <- tnrs_match_names(names = ("populus deltoides"))
populus #looks ok
species_Curtis1998 <- gsub("Populus euramericana", "Populus deltoides", species_Curtis1998) #replace populus euramericana with populus deltoides
```

@@

```{r dwonload one tree from treebase, results=FALSE}
Acers_tree <- search_treebase("418", by="id.tree", exact_match = TRUE) #returns a single tree in a list
Acers_tree <- Acers_tree[[1]] #extract tree from the list
```

@@

```{r plot pruned Acers_tree, fig.width=10, fig.height=4, echo=TRUE, message=FALSE}
plot(Acers4sp_tree, cex=.8, label.offset =.1, no.margin = TRUE)
```

@@

```{r check Acers_tree}
###check the overlap between species name from the data fila nad the tree file
intersect(as.character(Acers4sp_tree$tip.label), Acer4sp) # all matching
setdiff(Acer4sp, as.character(Acers4sp_tree$tip.label)) #no mismatches
is.binary.tree(tree_random) #TRUE
```

@@

Use *treebase* to create a tree for the three Quercus species from dat.Curtis1998.

>>>>

**Setup:** Work can be done individually, in pairs or small groups. R/Rstudio and Internet connection needed.

@@

A few things to note:  
1. We can use a tree with or without branch lengths (divergence measures).

```{r tree exmaple1, eval=TRUE, echo=FALSE, fig.width=6, fig.height=4, fig.cap=" "}
library(ape)
par(mfcol=c(1,2))
plot(read.tree(text = "(((Species4, Species3), Species2), Species1);"), main = "no branch lengths")
plot(read.tree(text = "(((Species4:0.1, Species3:0.1):0.3, Species2:0.4):1.9, Species1:2.3);"), main = "with branch lengths", use.edge.length = TRUE)
```

2. The tree has to be binary (strictly bifurcating = no polytomies, i.e. only two branches are coming from each node). Here is a simple example:

```{r tree exmaple2, eval=TRUE, echo=FALSE, fig.width=6, fig.height=4, fig.cap=" "}
par(mfcol=c(1,2))
plot(read.tree(text = "(((Species4, Species3), Species2), Species1);"), main = "binary")
plot(read.tree(text = "((Species4, Species3, Species2), Species1);"), main = "non-binary")
```

3. The tree needs to be rooted (has a node representing the most recent common ancestor of all the species in the tree).

```{r tree exmaple3, eval=TRUE, echo=FALSE, fig.width=6, fig.height=4, fig.cap=" "}
par(mfcol=c(1,2))
plot(read.tree(text = "(((Species4, Species3), Species2), Species1);"), main = "rooted")
plot(read.tree(text = "((Species4, Species3), Species2, Species1);"), main = "unrooted")
```

4. The tree tip labels (representing species) must match **exactly** with the species names in the data set (otherwise MA models will not run with them).

```{r tree exmaple4, eval=TRUE, echo=FALSE, fig.width=6, fig.height=4, fig.cap=" "}
par(mfcol=c(1,1))
plot(read.tree(text = "(((Species4 != SPECIES4, Species3 == Species3), Species2 == Species2), Species1 == Species1);"), main = "rooted")
```

5. Trees can be also made and used for higher taxonomic levels than species, e.g. genus-level (not covered in this tutorial, but general principles are the same).

```{r tree exmaple5, eval=TRUE, echo=FALSE, fig.width=6, fig.height=3.2, fig.cap=" "}
par(mfcol=c(1,1))
plot(read.tree(text = "((GenusC, GenusB), GenusA);"), main = "genus-level tree")
```

The general flow of tree-making work is shown in the diagram below.  
You will usually start with species list from the meta-analytic data set (use binomial Latin names for the species list and in your database). First, check if there is any phylogenetic tree available that already contains all your species (e.g. a super-tree). Currently, the easiest way to build a custom phylogenetic tree usable for meta-analysis is by using an R package called *rotl*, which accesses a synthetic super-tree from Open Tree of Life database (https://opentree.wikispaces.com/). We will use this approach in our examples below. Most of the time *rotl* will work well, especially with the common species. There are other super-trees available, e.g. for birds - we will use one of these as our example too. Overall, if you can find a tree that contains all your species, you just trim (subset) it down to your species. If a few species is missing, it is sometimes possible to substitute them with closely related species. If no existing phylogeny is readily usable, there are a few non-exclusive options depending on the range of species in the data set:  
 - If the species are very diverged (e.g. mixture of vertebrates and invertebrates), use NCBI Taxonomy Common Tree tool (http://www.ncbi.nlm.nih.gov/Taxonomy/CommonTree/wwwcmt.cgi). At the webpage, use an input box at the top of the page to search each species name or upload list of the species names from a file. This allows to generate a scaffold for a tree (i.e. the pattern of basal branching for the main clades).  
 - If few major clades are present, assemble together sub-trees created from bigger trees, if these exist for these clades.    
 - Use phyloGenerator (http://willpearse.github.io/phyloGenerator/) or other phylogenetic software to create phylogeny based on DNA sequences, if a shared sequence set exists for a given set of species (check if the resulting phylogeny makes sense).

@@

Run the below code chanks IF you have not installed these packages earlier (it may take 15-20 minutes to install all).

@@

Extract and clean the list of species included. Note that genus and species name are stored in separate columns. Also, smae species name can appear several times in the data set, but since we need a species list, we need each name to apper only once (i.e. unique names), and capitalisation to be consistent (letter case matters when matching strings).

@@

**Note:** populus deltoides

@@

So, we run into a problem with one of the Populus names - need to fix this by using an alternative name. Populus euramericana, is a naturally occurring hybrid of Populus deltoides, so we can replace it with this name for the search (given its colesely related and we dont have this species on our species list).

```{r fix Curtis1998 species_Curtis1998 list}
populus <- tnrs_match_names(names = ("populus canadensis"))
populus #looks ok
species_Curtis1998 <- gsub("Populus euramericana", "Populus deltoides", species_Curtis1998) #replace populus euramericana with populus canadensis
```

@@

```{r dwonload one tree from treebase, results=FALSE}
Acers_tree <- search_treebase("418", by="id.tree", exact_match = TRUE) #returns a single tree in a list
Acers_tree <- Acers_tree[[1]] #extract the tree from the list
```

@@

```{r plot pruned Acers_tree, fig.width=10, fig.height=3.2, echo=TRUE, message=FALSE}
plot(Acers4sp_tree, cex=.8, label.offset =.1, no.margin = TRUE)
```

@@

```{r check Acers_tree}
###check the overlap between species name from the data fila nad the tree file
intersect(as.character(Acers4sp_tree$tip.label), Acer4sp) # all matching
setdiff(Acer4sp, as.character(Acers4sp_tree$tip.label)) #no mismatches
is.binary.tree(Acers4sp_tree) #TRUE
```

We can graft this tree on the Curtis1998 tree to non-randomly resolve polytomy for genus Acer.   
**Note:** when grafting you lose branch lengths on the original tree, if present.

We first need to prune given branch to remove all but one species affected by polytomy. Then we can replace this species with the resolved sub-tree (Acers4sp_tree):

```{r graft Acers_tree}
tree_pruned <- drop.tip(tree, c("Acer saccharum", "Acer rubrum", "Acer saccharinum")) #prune 3 species
tree_grafted <- bind.tree(tree_pruned, Acers4sp_tree, where=which(tree_pruned$tip.label=="Acer pensylvanicum")) #graft 3 species to the remaining Acer species
```

```{r plot tree_grafted Acers, fig.width=10, fig.height=8, echo=TRUE, message=FALSE}
plot(tree_grafted, cex=.8, label.offset =.1, no.margin = TRUE)
```

@@

Use *treebase* to create a tree for the three Quercus species from dat.Curtis1998. You can try to graft them on the Curtis 1998 tree.
